---
title: "Does It Get the Right Answer? Validating IndepAssoc Against Known Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Does It Get the Right Answer? Validating IndepAssoc Against Known Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The simulated examples in the quick-start vignette prove the pipeline runs;
this one checks whether it gets the *right* answer. The logic is that of an
answer key: use datasets whose correct result is already known, run the
package's methods on them, and compare what comes out with what should be true.

The first dataset is NHEFS (NHANES Epidemiologic Follow-up Study), from the
`causaldata` package — the canonical textbook example for regression, matching,
stratification, IPTW, and AIPW (Hernán & Robins, *Causal Inference: What If*).
Its findings are well established, so agreement between the package's results
and the textbook answers is meaningful: the scientific question is whether
quitting smoking is independently associated with an outcome after adjusting
for measured confounders.

The second is the NSW demonstration, a randomized job-training program whose
effect on 1978 earnings was measured directly in the experiment — the Dehejia &
Wahba (1999) benchmark estimate of **$1,794**. `MatchIt::lalonde` combines the
program's treated units with a nonexperimental comparison group, so we can run
the package's observational methods on that combined data and check whether
they recover the answer the experiment already gave us.

## The five methods on NHEFS

```{r data}
nhefs_ok <- requireNamespace("causaldata", quietly = TRUE)
if (nhefs_ok) {
  nhefs <- causaldata::nhefs
  covs <- c("sex", "age", "race", "education", "smokeintensity",
            "smokeyrs", "exercise", "active", "wt71")
  nhefs_cc <- nhefs[!is.na(nhefs$wt82_71), ]
}
if (!nhefs_ok) {
  message("The `causaldata` package is not installed; ",
          "the analysis chunks below are skipped.")
}
```

The binary exposure is `qsmk` (quit smoking, 1 = yes) and the covariates are
the standard set used in the textbook worked example: age, sex, race,
education, smoking intensity, smoking years, exercise, activity, and baseline
weight.

### Binary outcome: death

```{r binary-pipeline, eval = nhefs_ok}
library(IndepAssoc)

# We set a fixed random seed (1) before running the pipeline so the matching
# step is reproducible; see `?run_pipeline`.
set.seed(1)
result_death <- run_pipeline(
  data = nhefs,
  exposure = "qsmk",
  covariates = covs,
  outcome = "death",
  type = "binary",
  methods = c("regression", "matching", "stratification", "iptw", "aipw")
)
format_comparison(result_death$comparison)
```

All five methods — outcome regression, matching, stratification, IPTW, and
AIPW — return odds ratios close to 1 (the value that means no association),
with confidence intervals that straddle 1 (they include that no-effect value)
and p-values well above 0.05. The agreement across five functionally different
adjustment strategies (those that model the outcome directly and those that
model treatment assignment) is reassuring: the null finding is not an artifact
of any single estimator. Had the estimates diverged sharply — say, matching far
from regression — that would flag outcome-model misspecification or poor
propensity-score overlap and warrant closer inspection of the
propensity-score distribution.

```{r binary-forest, eval = nhefs_ok}
plot_comparison(result_death$comparison, log_scale = TRUE)
```

This is a confounder-adjusted association under the standard
no-unmeasured-confounding assumption, not a proven causal effect.

### Continuous outcome: weight change (wt82_71)

Here `nhefs_cc` contains only the 1566 complete cases (the 63 rows with missing
`wt82_71` are excluded; the restriction is stated here for transparency).

```{r continuous-pipeline, eval = nhefs_ok}
# We set a fixed random seed (1) before running the pipeline so the matching
# step is reproducible; see `?run_pipeline`.
set.seed(1)
result_wt <- run_pipeline(
  data = nhefs_cc,
  exposure = "qsmk",
  covariates = covs,
  outcome = "wt82_71",
  type = "continuous",
  methods = c("regression", "matching", "stratification", "iptw", "aipw")
)
format_comparison(result_wt$comparison)
```

Here `estimate` is the adjusted mean weight change in kilograms associated with
quitting smoking. All five methods agree on a gain of roughly 3 to 3.5 kg, with
narrow confidence intervals that exclude 0 (so the gain is not plausibly
zero). The convergence across methods is consistent with the well-documented
finding from this dataset that quitting smoking is associated with weight gain.

```{r continuous-forest, eval = nhefs_ok}
plot_comparison(result_wt$comparison, log_scale = FALSE)
```

This is a confounder-adjusted association under the standard
no-unmeasured-confounding assumption, not a proven causal effect.

## The matching and weighting estimates against an experimental benchmark

```{r lalonde-data}
lalonde_ok <- requireNamespace("MatchIt", quietly = TRUE)
if (lalonde_ok) {
  lalonde <- MatchIt::lalonde
}
if (!lalonde_ok) {
  message("The `MatchIt` package is not installed; ",
          "the analysis chunks below are skipped.")
}
```

Observational studies cannot be validated against an experimental benchmark the
way randomized trials can — but a few classic datasets let us do exactly that.
`MatchIt::lalonde` combines the 185 experimental treated units with a
nonexperimental comparison group (the Dehejia–Wahba sample). Applying
observational methods to this hybrid dataset and comparing their estimates to
the known $1,794 benchmark is a standard sanity check: methods that land near
the benchmark are doing the adjustment job well on real, messy data. Here we
apply the two propensity-score methods the benchmark is designed to validate —
`matching` (1:1 nearest-neighbor PSM with a 0.2 caliper and within-pair
estimation) and `iptw` (stabilized inverse-probability weights) — to `re78`
(earnings in 1978, continuous outcome), with the standard covariate set.

```{r benchmark-pipeline, eval = lalonde_ok}
# We set a fixed random seed (1) before running the pipeline so the matching
# step is reproducible; see `?run_pipeline`.
set.seed(1)
result <- run_pipeline(
  data = lalonde,
  exposure = "treat",
  covariates = c("age", "educ", "race", "married", "nodegree", "re74", "re75"),
  outcome = "re78",
  type = "continuous",
  methods = c("matching", "iptw")
)
format_comparison(result$comparison)
```

Both estimates are the adjusted mean difference in 1978 earnings (dollars)
associated with the NSW program. The experimental benchmark is $1,794.

```{r benchmark-forest, eval = lalonde_ok}
plot_comparison(result$comparison, log_scale = FALSE)
```

The propensity-score matching estimate lands close to the experimental
benchmark of $1,794, while the IPTW estimate deviates further — a pattern that
mirrors the original Dehejia & Wahba finding that PSM recovers the experimental
ATT (the average treatment effect among the treated — the program's effect on
the people who actually took part) much better than naive or heavily weighted
estimators on this dataset. Divergence from the benchmark is itself
informative, signaling sensitivity to the weighting specification or limited
covariate overlap.

This is a confounder-adjusted association under the standard
no-unmeasured-confounding assumption, not a proven causal effect.

## References

- Hernán, M.A. and Robins, J.M. (2020). *Causal Inference: What If*. Chapman &
  Hall/CRC.
- Dehejia, R.H. and Wahba, S. (1999). Causal Effects in Nonexperimental
  Studies: Reevaluating the Evaluation of Training Programs. *Journal of the
  American Statistical Association*, 94(448), pp.1053-1062.
- Lalonde, R. (1986). Evaluating the Econometric Evaluations of Training
  Programs with Experimental Data. *American Economic Review*, 76(4), pp.604-620.
- The NHEFS data are distributed with the `causaldata` package (Nick Huntington-Klein and Malcolm Barrett); see `citation("causaldata")`.
- The `lalonde` data are distributed with the `MatchIt` package (Ho, Imai, King
  and Stuart); see `citation("MatchIt")`.
