---
title: "IndepAssoc: A Quick-Start Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IndepAssoc: A Quick-Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

`IndepAssoc` generalizes the propensity-score-based analysis pipeline used in
retrospective cohort studies to answer the core scientific question:

> Is a risk factor or exposure independently associated with an outcome, after
> adjusting for confounders?

Matching is used in the context of estimating the effect of a binary treatment
or exposure on an outcome while controlling for measured pre-treatment
variables. The goal of matching is to produce *covariate balance*, i.e., for
the distributions of covariates in the two groups to be approximately equal, as
they would be in a successful randomized experiment.

A matching analysis with `IndepAssoc` involves four primary steps: 1) building
a propensity score model, 2) matching cohorts, 3) assessing covariate balance,
and 4) estimating the treatment effect using multiple regression approaches.
Here we walk through these steps using a simulated example dataset.

```{r load-packages}
library(IndepAssoc)

data(example_cohort)
data <- example_cohort
head(data, 5)
```

The dataset contains a binary exposure indicator (`exposure`), baseline
covariates (`age`, `diabetes`, `hypertension`, `bmi`), and two outcomes
(`outcome_binary`, `outcome_continuous`).

## Step 1: Build the Propensity Score Model

The first step is to estimate a propensity score for each unit — the predicted
probability of receiving treatment given the covariates. This is done via
logistic regression.

```{r ps-model}
ps <- build_ps_model(
  data = data,
  exposure = "exposure",
  covariates = c("age", "diabetes", "hypertension", "bmi")
)

summary(ps$model)$coefficients
```

The propensity scores are appended to the data as the `.ps` column:

```{r ps-scores}
summary(ps$data$.ps)
```

## Step 2: Check Initial Imbalance

Before matching, we can examine the initial imbalance using
`cobalt::bal.tab()` via `check_balance()`:

```{r initial-balance, fig.width=8, fig.height=4}
m0 <- match_cohort(ps, caliper = 0.2)
check_balance(m0, threshold = 0.10, plot = FALSE)
```

## Step 3: Match Cohorts

We perform 1:1 nearest-neighbor matching on the propensity score without
replacement, with a caliper of 0.2 standard deviations of the logit(PS):

```{r match}
matched <- match_cohort(ps, caliper = 0.2)
cat("Unmatched:", nrow(data), "-> Matched:", nrow(matched$data), "observations\n")
```

The matched data includes a `match_num` (and `strata`) column identifying each
matched pair.

## Step 4: Assess Covariate Balance

After matching, we assess whether the covariates are balanced between treatment
groups:

```{r balance}
balance <- check_balance(matched, threshold = 0.10)
cat("All balanced:", balance$all_balanced, "\n")
```

### Love Plot

`run_pipeline()` returns the ASMD chart as part of its result. The
`balance_plot` element visualizes absolute standardized mean differences (ASMD)
for each covariate before and after matching:

```{r love-plot, fig.width=8, fig.height=5}
result <- run_pipeline(
  data = data,
  exposure = "exposure",
  covariates = c("age", "diabetes", "hypertension", "bmi"),
  outcome = "outcome_binary",
  type = "binary",
  methods = "regression"
)

result$balance_plot
```

Blue bars represent the unmatched cohort and orange bars represent the matched
cohort. All covariates should fall below the dashed line (ASMD < 0.10) after
matching.

## Step 5: Descriptive Tables

### Unmatched Data

For the unmatched data, we use chi-square tests (categorical) and Wilcoxon
rank-sum tests (continuous):

```{r table-unmatched}
table_unmatched(data, "exposure", c("age", "diabetes", "hypertension", "bmi"))
```

### Matched Data

For the matched data, we use paired tests — McNemar (categorical) and
Wilcoxon signed-rank (continuous):

```{r table-matched}
table_matched(matched, c("age", "diabetes", "hypertension", "bmi"))
```

## Step 6: Estimate the Treatment Effect

`IndepAssoc` fits three regression models matching common reporting standards in
the cardiac surgery literature (Benedetto et al., 2018):

### Binary Outcome

```{r models-binary}
models_bin <- fit_all_models(
  ps_model = ps,
  matched_data = matched$data,
  outcome = "outcome_binary",
  type = "binary"
)

models_bin_display <- models_bin$summary_w
models_bin_display$p <- ifelse(models_bin_display$p < 0.001, "<0.001",
                                sprintf("%.3f", models_bin_display$p))
models_bin_display
```

The three models are:

1. **Fully adjusted logistic**: GLM on the full unmatched data with all covariates
2. **Conditional logit**: `clogit()` on matched data, stratified by pair
3. **Mixed effect logistic**: `glmer()` on matched data with random intercept for pair

### Continuous Outcome

```{r models-continuous}
models_cont <- fit_all_models(
  ps_model = ps,
  matched_data = matched$data,
  outcome = "outcome_continuous",
  type = "continuous"
)

models_cont_display <- models_cont$summary_w
models_cont_display$p <- ifelse(models_cont_display$p < 0.001, "<0.001",
                                 sprintf("%.3f", models_cont_display$p))
models_cont_display
```

### Compare All Confounding-Adjustment Methods

Beyond the three regression models above, `fit_outcome()` tests the exposure–outcome
association under five confounding-adjustment methods that share a common tidy
result: `regression` (full-sample adjusted GLM/LM), `matching` (propensity-score
matching with conditional-logit (binary) / within-pair (continuous) estimation),
`stratification` (Mantel–Haenszel quintiles for binary,
inverse-variance pooling for continuous), `iptw` (stabilized inverse-probability
weights with robust `sandwich` (`vcovHC`) standard errors), and `aipw` (Bang & Robins
augmented estimator). `run_pipeline(..., methods = ...)` runs them all and
returns a single comparison table; `format_comparison()` renders it as a
publication-ready table:

```{r pipeline-comparison}
result <- run_pipeline(
  data = data,
  exposure = "exposure",
  covariates = c("age", "diabetes", "hypertension", "bmi"),
  outcome = "outcome_binary",
  type = "binary",
  methods = c("regression", "matching", "stratification", "iptw", "aipw")
)

format_comparison(result$comparison)
```

All methods recover a positive association between treatment and the binary
outcome (`estimate` > 1 on the odds-ratio scale). Agreement across the five
methods — including the matched and weighted estimators that do not rely on the
outcome-regression functional form — is a reassuring sign that the association
is not an artifact of any single adjustment strategy. Large discrepancies
between, say, the regression and matching estimates often indicate
misspecification of the outcome model or poor overlap, and warrant closer
inspection of the propensity-score distribution.

The comparison table can be visualized directly:

```{r pipeline-forest}
plot_comparison(result$comparison)
```

## Choosing an adjustment method

The five methods answer the same question — is the exposure independently
associated with the outcome after adjusting for the measured confounders? —
but reach the answer by different routes, and the right choice depends on the
data and the question:

- **Regression** models the outcome directly with all confounders included.
  Efficient and familiar, but the estimate depends on the outcome-model
  functional form, and it extrapolates where propensity scores have poor
  overlap.
- **Matching** discards poorly overlaid units, so inference stays close to the
  data; it answers the treatment-on-the-treated question and is the estimator
  used in the source cardiac-surgery papers. Cost: fewer units and no benefit
  from observations that do not overlap.
- **Stratification** groups patients into propensity-score strata (quintiles)
  and pools within-stratum contrasts. Simple and transparent, but residual
  confounding within strata is a real risk when the score's discriminating
  power is high.
- **IPTW** reweights the sample to mimic a randomized comparison. It uses all
  units, but very small weights on poorly overlaid patients can inflate
  variance and invite model sensitivity.
- **AIPW** combines an outcome model with a treatment model so the estimate
  stays consistent if only one of the two is correctly specified. Doubly
  robust, and the natural choice when both models are plausible.

In practice: run all five through `run_pipeline()` and read the comparison
table. Agreement across methods that model the outcome directly (regression,
AIPW) and those that model treatment assignment (matching, stratification,
IPTW) is evidence the association is not an artifact of any single adjustment
strategy. Disagreement flags misspecification or poor overlap, not a vote —
inspect the propensity-score distribution and the balance tables before
trusting any single estimate.

## Step 7: Paired Statistical Tests

### McNemar Test (Binary)

```{r mcnemar}
mcnemar_result <- mcnemar_test(matched$data, "outcome_binary", "exposure")
mcnemar_result$p.value <- sprintf("%.3f", mcnemar_result$p.value)
mcnemar_result$statistic <- round(mcnemar_result$statistic, 2)
mcnemar_result$p <- sprintf("%.3f", mcnemar_result$p)
mcnemar_result

```

### Paired Wilcoxon Test (Continuous)

```{r wilcoxon}
wilcox_result <- paired_wilcoxon_test(matched$data, "outcome_continuous", "exposure")
wilcox_result$p.value <- sprintf("%.3f", wilcox_result$p.value)
wilcox_result
```

## Full Pipeline

All steps above can be executed in a single call:

```{r pipeline, eval=FALSE}
result <- run_pipeline(
  data = data,
  exposure = "exposure",
  covariates = c("age", "diabetes", "hypertension", "bmi"),
  outcome = "outcome_binary",
  type = "binary"
)

summary(result)
```

## Exporting Results

```{r export, eval=FALSE}
export_results(result, output_dir = "output")
```

This writes CSV files for the balance tables, regression summaries, and
statistical tests.

## Reporting Results

To report matching results in a manuscript, include:

1. The matching specification (method, caliper, ratio)
2. The distance measure (logistic regression PS)
3. Covariate balance statistics (ASMD before and after matching, Love plot)
4. Number of matched, unmatched, and discarded units
5. The method of estimating the treatment effect and standard error
6. All three regression model results (fully adjusted, conditional,
   mixed effect)

### Example write-up

> We used propensity score matching to estimate the association between
> treatment and outcome while adjusting for age, diabetes, hypertension, and
> BMI. A logistic regression model was used to estimate propensity
> scores, and 1:1 nearest-neighbor matching without replacement was performed
> with a caliper of 0.2 standard deviations of the logit propensity score.
> After matching, all standardized mean differences for covariates were below
> 0.10, indicating adequate balance (Figure 1). Three regression models were
> fitted to the matched cohort: (1) a fully adjusted logistic regression on
> the unmatched data, (2) a conditional logistic regression on the matched
> data stratified by pair, and (3) a mixed-effect logistic regression
> with a random intercept for matched pair. All three models showed a
> statistically significant association between treatment and outcome (Table 1).

## References

- Benedetto, U., Head, S.J., Angelini, G.D. and Blackstone, E.H. (2018).
  Statistical primer: propensity score matching and its alternatives.
  *European Journal of Cardio-Thoracic Surgery*, 53(6), pp.1112-1117.

- Ho, D.E., Imai, K., King, G. and Stuart, E.A. (2007). Matching as
  Nonparametric Preprocessing for Reducing Model Dependence in Parametric
  Causal Inference. *Political Analysis*, 15(3), pp.199-236.

- Rosenbaum, P.R. and Rubin, D.B. (1983). The Central Role of the Propensity
  Score in Observational Studies for Causal Effects. *Biometrika*, 70(1),
  pp.41-55.

- Stuart, E.A. (2010). Matching Methods for Causal Inference: A Review and a
  Look Forward. *Statistical Science*, 25(1), pp.1-21.
