---
title: "IndepAssoc: Confounder-Adjusted Association Testing via Propensity Score Methods"
author: "Mostafa Elshamy"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 7
    fig_height: 5
    keep_tex: true
vignette: >
  %\VignetteIndexEntry{IndepAssoc: Confounder-Adjusted Association Testing via Propensity Score Methods (PDF)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

`IndepAssoc` generalizes the propensity-score-based analysis pipeline used in
retrospective cohort studies to answer the core scientific question:

> *Is a risk factor or exposure independently associated with an outcome, after
> adjusting for confounders?*

Matching is used in the context of estimating the effect of a binary treatment
or exposure on an outcome while controlling for measured pre-treatment
variables. The goal of matching is to produce **covariate balance**, i.e., for
the distributions of covariates in the two groups to be approximately equal, as
they would be in a successful randomized experiment.

A matching analysis with `IndepAssoc` involves four primary steps:

1. Building a propensity score model
2. Matching cohorts
3. Assessing covariate balance
4. Estimating the treatment effect using multiple regression approaches

```{r load-packages}
library(IndepAssoc)

data <- read.csv(system.file("extdata/sample_psm_dataset.csv", package = "IndepAssoc"))
head(data, 5)
```

\newpage

# Propensity Score Model

The first step is to estimate a propensity score for each unit---the predicted
probability of receiving treatment given the covariates, estimated via logistic
regression.

```{r ps-model}
ps <- build_ps_model(
  data = data,
  exposure = "treatment",
  covariates = c("age", "bmi", "smoker", "hypertension")
)

knitr::kable(
  summary(ps$model)$coefficients,
  caption = "Propensity score model coefficients",
  digits = 4
)
```

\newpage

# Matching

We perform 1:1 nearest-neighbor matching on the propensity score without
replacement, with a caliper of 0.2 standard deviations of the logit(PS).

```{r match}
matched <- match_cohort(ps, caliper = 0.2)
cat("Unmatched:", nrow(data), "-> Matched:", nrow(matched$data), "observations\n")
```

\newpage

# Covariate Balance

```{r balance}
balance <- check_balance(matched, threshold = 0.10)
cat("All balanced:", balance$all_balanced, "\n")
```

## Love Plot

```{r love-plot, fig.cap="Love plot: ASMD before and after matching.", fig.width=7, fig.height=4}
plot_love(matched, threshold = 0.10)
```

\newpage

# Descriptive Tables

## Unmatched Data

```{r table-unmatched}
table_unmatched(data, "treatment", c("age", "bmi", "smoker", "hypertension"))
```

## Matched Data

```{r table-matched}
table_matched(matched, c("age", "bmi", "smoker", "hypertension"))
```

\newpage

# Outcome Models

## Binary Outcome

```{r models-binary}
models_bin <- fit_all_models(
  ps_model = ps,
  matched_data = matched$data,
  outcome = "binary_outcome",
  type = "binary"
)

knitr::kable(models_bin$summary_w, caption = "Binary outcome models", digits = 4)
```

## Continuous Outcome

```{r models-continuous}
models_cont <- fit_all_models(
  ps_model = ps,
  matched_data = matched$data,
  outcome = "numeric_outcome",
  type = "continuous"
)

knitr::kable(models_cont$summary_w, caption = "Continuous outcome models", digits = 4)
```

\newpage

# Paired Statistical Tests

## McNemar Test

```{r mcnemar}
mcnemar_test(matched$data, "binary_outcome", "treatment")
```

## Paired Wilcoxon Test

```{r wilcoxon}
paired_wilcoxon_test(matched$data, "numeric_outcome", "treatment")
```

\newpage

# Full Pipeline

```{r pipeline, eval=FALSE}
result <- run_pipeline(
  data = data,
  exposure = "treatment",
  covariates = c("age", "bmi", "smoker", "hypertension"),
  outcome = "binary_outcome",
  type = "binary"
)
summary(result)
```

# Exporting Results

```{r export, eval=FALSE}
export_results(result, output_dir = "output")
```

\newpage

# References

- Benedetto, U., Head, S.J., Angelini, G.D. and Blackstone, E.H. (2018).
  Statistical primer: propensity score matching and its alternatives.
  *European Journal of Cardio-Thoracic Surgery*, 53(6), pp.1112-1117.

- Ho, D.E., Imai, K., King, G. and Stuart, E.A. (2007). Matching as
  Nonparametric Preprocessing for Reducing Model Dependence in Parametric
  Causal Inference. *Political Analysis*, 15(3), pp.199-236.

- Rosenbaum, P.R. and Rubin, D.B. (1983). The Central Role of the Propensity
  Score in Observational Studies for Causal Effects. *Biometrika*, 70(1),
  pp.41-55.

- Stuart, E.A. (2010). Matching Methods for Causal Inference: A Review and a
  Look Forward. *Statistical Science*, 25(1), pp.1-21.
