---
title: "IndepAssoc: Confounder-Adjusted Association Testing via Propensity Score Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IndepAssoc: Confounder-Adjusted Association Testing via Propensity Score Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

`IndepAssoc` generalizes the propensity-score-based analysis pipeline used in
retrospective cohort studies to answer the core scientific question:

> Is a risk factor or exposure independently associated with an outcome, after
> adjusting for confounders?

Matching is used in the context of estimating the effect of a binary treatment
or exposure on an outcome while controlling for measured pre-treatment
variables. The goal of matching is to produce *covariate balance*, i.e., for
the distributions of covariates in the two groups to be approximately equal, as
they would be in a successful randomized experiment.

A matching analysis with `IndepAssoc` involves four primary steps: 1) building
a propensity score model, 2) matching cohorts, 3) assessing covariate balance,
and 4) estimating the treatment effect using multiple regression approaches.
Here we walk through these steps using a simulated example dataset.

```{r load-packages}
library(IndepAssoc)

data(example_cohort)
data <- example_cohort
head(data, 5)
```

The dataset contains a binary exposure indicator (`exposure`), baseline
covariates (`age`, `diabetes`, `hypertension`, `bmi`), and two outcomes
(`outcome_binary`, `outcome_continuous`).

## Step 1: Build the Propensity Score Model

The first step is to estimate a propensity score for each unit — the predicted
probability of receiving treatment given the covariates. This is done via
logistic regression.

```{r ps-model}
ps <- build_ps_model(
  data = data,
  exposure = "exposure",
  covariates = c("age", "diabetes", "hypertension", "bmi")
)

summary(ps$model)$coefficients
```

The propensity scores are appended to the data as the `.ps` column:

```{r ps-scores}
summary(ps$data$.ps)
```

## Step 2: Check Initial Imbalance

Before matching, we can examine the initial imbalance using
`cobalt::bal.tab()` via `check_balance()`:

```{r initial-balance, fig.width=8, fig.height=4}
m0 <- match_cohort(ps, caliper = 0.2)
check_balance(m0, threshold = 0.10, plot = FALSE)
```

## Step 3: Match Cohorts

We perform 1:1 nearest-neighbor matching on the propensity score without
replacement, with a caliper of 0.2 standard deviations of the logit(PS):

```{r match}
matched <- match_cohort(ps, caliper = 0.2)
cat("Unmatched:", nrow(data), "-> Matched:", nrow(matched$data), "observations\n")
```

The matched data includes a `match_num` (and `strata`) column identifying each
matched pair.

## Step 4: Assess Covariate Balance

After matching, we assess whether the covariates are balanced between treatment
groups:

```{r balance}
balance <- check_balance(matched, threshold = 0.10)
cat("All balanced:", balance$all_balanced, "\n")
```

### Love Plot

The Love plot visualizes absolute standardized mean differences (ASMD) for each
covariate before and after matching:

```{r love-plot, fig.width=8, fig.height=5}
plot_love(matched, threshold = 0.10)
```

Blue bars represent the unmatched cohort and orange bars represent the matched
cohort. All covariates should fall below the dashed line (ASMD < 0.10) after
matching.

## Step 5: Descriptive Tables

### Unmatched Data

For the unmatched data, we use chi-square tests (categorical) and Wilcoxon
rank-sum tests (continuous):

```{r table-unmatched}
table_unmatched(data, "exposure", c("age", "diabetes", "hypertension", "bmi"))
```

### Matched Data

For the matched data, we use paired tests — McNemar (categorical) and
Wilcoxon signed-rank (continuous):

```{r table-matched}
table_matched(matched, c("age", "diabetes", "hypertension", "bmi"))
```

## Step 6: Estimate the Treatment Effect

`IndepAssoc` fits three regression models matching common reporting standards in
the cardiac surgery literature (Benedetto et al., 2018):

### Binary Outcome

```{r models-binary}
models_bin <- fit_all_models(
  ps_model = ps,
  matched_data = matched$data,
  outcome = "outcome_binary",
  type = "binary"
)

models_bin$summary_w
```

The three models are:

1. **Fully adjusted logistic**: GLM on the full unmatched data with all covariates
2. **Conditional logit**: `clogit()` on matched data, stratified by pair
3. **Mixed effect logistic**: `glmer()` on matched data with random intercept for pair

### Continuous Outcome

```{r models-continuous}
models_cont <- fit_all_models(
  ps_model = ps,
  matched_data = matched$data,
  outcome = "outcome_continuous",
  type = "continuous"
)

models_cont$summary_w
```

### Compare All Confounding-Adjustment Methods

Beyond the three regression models above, `fit_outcome()` tests the exposure–outcome
association under five confounding-adjustment methods that share a common tidy
result: `regression` (full-sample adjusted GLM/LM), `matching` (propensity-score
matching with conditional-logit (binary) / within-pair (continuous) estimation),
`stratification` (Mantel–Haenszel quintiles for binary,
inverse-variance pooling for continuous), `iptw` (stabilized inverse-probability
weights with robust `sandwich` (`vcovHC`) standard errors), and `aipw` (Bang & Robins
augmented estimator). `run_pipeline(..., methods = ...)` runs them all and
returns a single comparison table:

```{r pipeline-comparison}
result <- run_pipeline(
  data = data,
  exposure = "exposure",
  covariates = c("age", "diabetes", "hypertension", "bmi"),
  outcome = "outcome_binary",
  type = "binary",
  methods = c("regression", "matching", "stratification", "iptw", "aipw")
)

result$comparison
```

All methods recover a positive association between treatment and the binary
outcome (`estimate` > 1 on the odds-ratio scale). Agreement across the five
methods — including the matched and weighted estimators that do not rely on the
outcome-regression functional form — is a reassuring sign that the association
is not an artifact of any single adjustment strategy. Large discrepancies
between, say, the regression and matching estimates often indicate
misspecification of the outcome model or poor overlap, and warrant closer
inspection of the propensity-score distribution.

The comparison table can be visualized directly:

```{r pipeline-forest}
plot_comparison(result$comparison)
```

## Step 7: Paired Statistical Tests

### McNemar Test (Binary)

```{r mcnemar}
mcnemar_test(matched$data, "outcome_binary", "exposure")
```

### Paired Wilcoxon Test (Continuous)

```{r wilcoxon}
paired_wilcoxon_test(matched$data, "outcome_continuous", "exposure")
```

## Full Pipeline

All steps above can be executed in a single call:

```{r pipeline, eval=FALSE}
result <- run_pipeline(
  data = data,
  exposure = "exposure",
  covariates = c("age", "diabetes", "hypertension", "bmi"),
  outcome = "outcome_binary",
  type = "binary"
)

summary(result)
```

## Exporting Results

```{r export, eval=FALSE}
export_results(result, output_dir = "output")
```

This writes CSV files for the balance tables, regression summaries, and
statistical tests.

## Reporting Results

To report matching results in a manuscript, include:

1. The matching specification (method, caliper, ratio)
2. The distance measure (logistic regression PS)
3. Covariate balance statistics (ASMD before and after matching, Love plot)
4. Number of matched, unmatched, and discarded units
5. The method of estimating the treatment effect and standard error
6. All three regression model results (fully adjusted, conditional,
   mixed effect)

### Example write-up

> We used propensity score matching to estimate the association between
> treatment and outcome while adjusting for age, diabetes, hypertension, and
> BMI. A logistic regression model was used to estimate propensity
> scores, and 1:1 nearest-neighbor matching without replacement was performed
> with a caliper of 0.2 standard deviations of the logit propensity score.
> After matching, all standardized mean differences for covariates were below
> 0.10, indicating adequate balance (Figure 1). Three regression models were
> fitted to the matched cohort: (1) a fully adjusted logistic regression on
> the unmatched data, (2) a conditional logistic regression on the matched
> data stratified by pair, and (3) a mixed-effect logistic regression
> with a random intercept for matched pair. All three models showed a
> statistically significant association between treatment and outcome (Table 1).

## References

- Benedetto, U., Head, S.J., Angelini, G.D. and Blackstone, E.H. (2018).
  Statistical primer: propensity score matching and its alternatives.
  *European Journal of Cardio-Thoracic Surgery*, 53(6), pp.1112-1117.

- Ho, D.E., Imai, K., King, G. and Stuart, E.A. (2007). Matching as
  Nonparametric Preprocessing for Reducing Model Dependence in Parametric
  Causal Inference. *Political Analysis*, 15(3), pp.199-236.

- Rosenbaum, P.R. and Rubin, D.B. (1983). The Central Role of the Propensity
  Score in Observational Studies for Causal Effects. *Biometrika*, 70(1),
  pp.41-55.

- Stuart, E.A. (2010). Matching Methods for Causal Inference: A Review and a
  Look Forward. *Statistical Science*, 25(1), pp.1-21.
