---
title: "Validating the Methods on the RHC Benchmark Cohort"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating the Methods on the RHC Benchmark Cohort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
library(IndepAssoc)
```

## Introduction

This vignette validates the package's five confounding-adjustment methods on a
real, well-known observational dataset: the Right Heart Catheterization (RHC)
cohort from the SUPPORT study. Connors et al. (1996, *JAMA*) enrolled 5,735
critically ill adults across five U.S. teaching hospitals (1989-1994) and
found that right heart catheterization — a diagnostic procedure, not a
treatment in itself — was associated with *increased* mortality and *greater*
resource use, contradicting the clinical assumption at the time that it was
beneficial. Because the direction of that finding is well established and has
been reproduced by dozens of independent reanalyses since, RHC is one of the
standard benchmark datasets in the causal-inference literature, and it gives
this package a real cohort to validate against, not just simulated data.

We analyze three outcomes: 30-day mortality (`dth30`), longer-term mortality
by 180 days (`death`), and hospital length of stay in days (`los`). We check
whether all five adjustment methods — two that model the outcome directly
(regression, matching) and three that model the propensity of receiving RHC
(stratification, IPTW, AIPW) — agree on the *direction* of the association.
Agreement across five methods that make different, largely independent
assumptions is evidence the finding isn't an artifact of any single model's
specification. It is not, by itself, proof of a causal effect — see the
Caveat section at the end.

### The five methods

- **Regression** — a single covariate-adjusted regression (logistic for
  binary outcomes, linear for continuous) fit on the full, unmatched cohort.
  Uses every patient, but its validity rests entirely on the outcome model
  being correctly specified; no propensity score is involved.
- **Matching** — 1:1 nearest-neighbor propensity-score matching (caliper 0.2
  SD of logit-PS), followed by conditional logistic regression (binary
  outcomes) or a pair-stratified linear model (continuous outcomes) on just
  the matched subset.
- **Stratification** — splits the cohort into propensity-score strata and
  pools the within-stratum associations into one overall estimate. Keeps the
  full sample and controls confounding by comparing only patients with
  similar propensity scores.
- **IPTW** — weights each patient by the inverse of their estimated
  probability of receiving the exposure they actually received, then fits an
  exposure-only weighted regression. As of `IndepAssoc` 0.4.0 this is a plain
  marginal structural model — the weights alone carry the adjustment, there
  is no additional covariate adjustment layered on top.
- **AIPW** — combines a propensity-score model and an outcome regression
  model via the augmented-IPW formula. It is "doubly robust": the estimate
  stays consistent if *either* model is correctly specified, not necessarily
  both, making it more resilient to model misspecification than regression
  or IPTW used alone.

### Estimand: what each method actually targets

The five methods do not all estimate the same causal quantity. Matching targets
the **average treatment effect on the treated (ATT)** — only treated patients
that found a control match remain in the analysis, so the estimate describes the
treated population by construction. IPTW and AIPW target the **average
treatment effect (ATE)** over the full analytic sample: every patient
contributes, reweighted (IPTW) or via augmented influence functions (AIPW), so
the comparison mimics two exposure groups drawn from the full sample's covariate
distribution. Stratification pools stratum-specific effects across the full
analytic sample — a common odds ratio via the Cochran-Mantel-Haenszel method
for binary outcomes, an inverse-variance-weighted mean difference for continuous
outcomes — a full-sample, stratum-pooled estimate rather than a strict marginal
ATE. (With the new `estimand = "ATT"` option, stratification instead pools each
stratum's effect weighted by its number of treated units, and is then genuinely
ATT-consistent.) Regression reports a **conditional** effect: the coefficient is
adjusted
for the listed covariates, and for binary outcomes the odds ratio is
*non-collapsible* (a mathematical property, not a modeling choice), so it is not
numerically equal to a marginal ATE even when the model is correct.

So when the five rows below agree in direction and rough magnitude, that is real
evidence of robustness — but they are five related quantities, not five
independent measurements of a single parameter.

## Data preparation

```{r load-data}
data(rhc_sample)
rhc <- rhc_sample$data
covariates <- rhc_sample$covariates
exposure <- "swang1"

nrow(rhc)
length(covariates)
table(RHC = rhc[[exposure]])
```

The cohort ships with the package as `rhc_sample`, already cleaned by
`prepare_rhc_data()`: the exposure `swang1` is coded 0/1 (1 = received RHC
within 24 hours of ICU admission), and the covariate vector holds 50 baseline
characteristics — demographics, primary/secondary disease category,
comorbidities, and physiologic measurements — chosen because they retain the
full sample under complete-case analysis. (Two variables from the raw file,
`adld3p` and `urin1`, are excluded: both are missing in more than half the
cohort, and including either would collapse the usable sample from 5,735
patients to a few hundred.)

### Outcomes

```{r outcomes-def}
outcomes <- data.frame(
  name  = c("dth30", "death", "los"),
  label = c("30-day mortality", "180-day mortality",
            "Hospital length of stay (days)"),
  type  = c("binary", "binary", "continuous"),
  stringsAsFactors = FALSE
)
```

`dth30` and `death` are not the same outcome, and the difference matters:
`dth30` is mortality within a fixed 30-day window; `death` captures mortality
over the study's longer, 180-day follow-up, so every patient with `dth30 = 1`
also has `death = 1`, but not the reverse. Reporting both, rather than
treating them as interchangeable, is deliberate — it lets us see whether an
early-mortality signal persists over a longer horizon.

A binary 0/1 outcome doesn't have a meaningful median or quartiles, so
summarizing it with `summary()` (as you would for a continuous variable)
produces numbers that are technically correct but not clinically
interpretable — it reports what looks like a five-number summary of "0.35 of
a death." The block below reports what actually matters for a binary
outcome — the event count and proportion — and reserves `summary()` for the
one genuinely continuous outcome, `los`. It also reports how much missing
data each outcome has, since that determines each outcome's analytic sample
size later in this vignette, rather than assuming it up front.

```{r outcome-summaries}
for (i in seq_len(nrow(outcomes))) {
  nm <- outcomes$name[i]
  cat("\n", outcomes$label[i], " (`", nm, "`):\n", sep = "")

  if (outcomes$type[i] == "binary") {
    tab <- table(rhc[[nm]], useNA = "ifany")
    print(tab)
    cat("Event proportion:",
        round(mean(rhc[[nm]] == 1, na.rm = TRUE), 3), "\n")
  } else {
    print(summary(rhc[[nm]]))
  }

  cat("Missing:", sum(is.na(rhc[[nm]])), "of", nrow(rhc), "\n")
}
```

## Running the pipeline

Each outcome is analyzed on its own complete-case sample — patients with
non-missing values for the exposure, all 50 covariates, and that specific
outcome. In practice `dth30` and `death` are fully observed on this cohort
(confirmed by the "Missing: 0" lines above), so their complete-case samples
are identical to the full 5,735-patient cohort; `los` has a small number of
patients missing a discharge or admission date, so its analytic sample is
marginally smaller. Filtering explicitly, rather than letting each internal
model silently drop missing rows on its own, makes the exact analytic sample
size for each outcome visible up front.

We pass `seed = 1` directly to `run_pipeline()` for every outcome, rather
than calling `set.seed()` once before the loop. `run_pipeline()`'s `seed`
argument resets the random state at the very start of each call, so — because
matching depends only on the exposure and covariates, never on the outcome
being analyzed — `dth30` and `death` end up drawing the *same* matched cohort
(they share an identical complete-case sample), while `los`'s matched cohort
can differ very slightly, since it starts from a marginally different set of
patients.

```{r run-pipeline}
methods <- c("regression", "matching", "stratification", "iptw", "aipw")

run_one_outcome <- function(name, type) {
  analytic_sample <- rhc[stats::complete.cases(rhc[, c(exposure, covariates, name)]), ]
  run_pipeline(
    data       = analytic_sample,
    exposure   = exposure,
    covariates = covariates,
    outcome    = name,
    type       = type,
    methods    = methods,
    seed       = 1
  )
}

results_list <- setNames(
  lapply(seq_len(nrow(outcomes)),
         function(i) run_one_outcome(outcomes$name[i], outcomes$type[i])),
  outcomes$name
)
```

## Balance check before and after matching

Matching depends only on the exposure and the 50 covariates, so the same
balance chart applies to every outcome that shares this cohort's exposure and
covariate set; we show it once, from the `dth30` run.

```{r balance-plot}
results_list$dth30$balance_plot
```

The dashed line marks the 0.10 absolute-standardized-mean-difference (ASMD)
threshold conventionally used to define adequate covariate balance. Most
covariates in this cohort start well above that line before matching — the
RHC literature has repeatedly documented substantial baseline imbalance
between patients who did and didn't receive RHC — and matching should bring
most, though not necessarily every, covariate under the threshold. Rather
than asserting which covariates end up on which side of that line, treat the
chart above as the answer: it reflects this specific run, not a general
claim.

### Truncated view: the 25 most-imbalanced covariates

The full chart above is dense by construction: the 50 covariates include
several multi-level categorical variables (`cat1`, `cat2`, `ninsclas`,
`income`, `race`), which expand to roughly 76 bars and leave the x-axis
labels hard to read. `plot_asmd_balance()` offers an opt-in `top_n` argument
that shows only the covariates with the largest unadjusted (pre-matching)
ASMD — which is what the chart exists to communicate — and labels the chart
with exactly how many of the total it is showing:

```{r balance-plot-top-n}
plot_asmd_balance(results_list$dth30, top_n = 25)
```

## Results by outcome

```{r results-loop, results = "asis"}
for (i in seq_len(nrow(outcomes))) {
  nm   <- outcomes$name[i]
  type <- outcomes$type[i]

  cat("\n\n###", outcomes$label[i], "\n\n")

  print(knitr::kable(
    format_comparison(results_list[[nm]]$comparison),
    align = "lrrrr"
  ))
  cat("\n\n")

  print(plot_comparison(results_list[[nm]]$comparison,
                         log_scale = (type == "binary")))
  cat("\n\n")
}
```

`dth30` and `death` report an odds ratio; `los` reports a mean difference in
days. `IndepAssoc`'s own validated testing of this exact cohort (the
package's original single-outcome RHC analysis) found all five methods
agreeing that RHC is associated with *increased* 30-day mortality — odds
ratios roughly in the 1.3-1.5 range, confidence intervals excluding 1 — and
with a *longer* hospital stay, roughly 2.5-3 additional days, confidence
intervals excluding 0. The `dth30` and `los` tables above should reproduce
that same pattern; if they don't, that's worth investigating rather than
dismissing, since it would mean something about the analytic sample or
package version differs from what was previously validated.

One caveat on `los` specifically: hospital length of stay is heavily
right-skewed and bounded at zero (in the summary above, median 14 days, mean
21.5, maximum 342). The linear model reports a mean difference on the original
day scale, which is a common simplification — including in the two source
papers this package generalizes — but not the most distributionally appropriate
choice for an outcome with this shape: a log-transform or a Gamma GLM would fit
it better. Read the `los` mean difference as a convenient, interpretable
summary, not as evidence that a day-scale linear model is the ideal
specification.

For `death`, there is no prior validated result to compare against in this
vignette — read the table above on its own terms. Given that every `dth30 =
1` patient is also `death = 1`, a similar direction of association would not
be surprising, but the 180-day window gives more time for events unrelated to
the index RHC decision to occur, so the magnitude need not match the 30-day
analysis exactly. The table is the answer; this paragraph is only a guide to
reading it, not a substitute for it.

## Combined comparison across outcomes

```{r combined-table}
combined_comparison <- do.call(rbind, lapply(seq_len(nrow(outcomes)), function(i) {
  nm <- outcomes$name[i]
  cbind(Outcome = outcomes$label[i],
        results_list[[nm]]$comparison)
}))

combined_formatted <- format_combined(combined_comparison)
knitr::kable(combined_formatted, align = "llrrrr", row.names = FALSE)
```

The two binary outcomes report an odds ratio and the continuous outcome
reports a mean difference in days — read the "Estimate" column together with
the "Outcome" column, not as one directly comparable number across all
fifteen rows.

## A note on matching sample size

The `matching` row's `n` in every table above is far smaller than the other
four methods' — this is expected, not an error. Only 2,184 of the 5,735
patients received RHC, so 1:1 matching can never retain more than roughly
twice that many patients even under a perfect match rate; the other four
methods use the full analytic sample because they either weight or model the
outcome directly rather than discarding unmatched patients. A smaller,
matching-based sample generally means wider confidence intervals for that row
relative to the others — worth keeping in mind before treating any one row's
significance (or lack of it) as more authoritative than another's.

## Propensity-score overlap and weight diagnostics

The balance chart above checks covariates, but the propensity-score methods
rest on a second assumption: **positivity** — every patient must have a
nonzero (and ideally non-extreme) chance of being in either group. Extreme
propensity scores near 0 or 1, or the extreme inverse-probability weights they
produce, can make IPTW/AIPW unstable even when covariate balance looks fine.
`check_positivity()` inspects both, using the same propensity-score model the
pipeline already built:

```{r positivity-check}
pos_check <- check_positivity(results_list$dth30$ps_model)
print(pos_check)
```

Two things stand out. First, the treated and control propensity-score
distributions overlap broadly (roughly 0.019-0.985 treated vs 0.003-0.955
control), so the cohort does not have a structural positivity violation — a reader
comparing only the two support ranges would conclude overlap is fine. Second,
`check_positivity()` still flags 12 of 5,735 units outside the default
[0.01, 0.99] support window, and the stabilized IPTW weights run from 0.39 to
20.2 — a 52-fold max-to-min ratio. That is a mild warning, not a failure: 12
extreme-tail patients out of 5,735 will not flip the analysis, but they are
exactly the kind of extremity that can inflate IPTW/AIPW variance in a smaller
sample — and the problem the `trim` argument exists for (see below). Every
`run_pipeline()` call now prints this same summary and returns it as
`result$positivity`, so the check is visible by default rather than something
to remember to run.

## Sensitivity checks on `los`

The main results report `los` as a day-scale mean difference from the
full-sample analysis. Three specification choices deserve a closer look: which
estimand the propensity-score methods target, how the estimate responds to
weight trimming, and whether the outcome's right-skew changes the conclusion
under a log transform.

### Estimand: ATE vs ATT

```{r los-estimand}
los_sample <- rhc[stats::complete.cases(rhc[, c(exposure, covariates, "los")]), ]

los_reg <- fit_outcome(los_sample, exposure, covariates, "los",
                       type = "continuous", method = "regression")
los_ate <- fit_outcome(los_sample, exposure, covariates, "los",
                       type = "continuous", method = c("iptw", "aipw"))
los_att <- fit_outcome(los_sample, exposure, covariates, "los",
                       type = "continuous", method = c("iptw", "aipw"),
                       estimand = "ATT")

row_from_fit <- function(outcome_lab, fit) {
  data.frame(Outcome = outcome_lab, method = fit$method, type = fit$type,
             estimate = fit$estimate, conf_low = fit$conf_low,
             conf_high = fit$conf_high, p_value = fit$p_value, n = fit$n)
}
estimand_rows <- rbind(
  row_from_fit("Regression", los_reg),
  row_from_fit("ATE", los_ate$iptw), row_from_fit("ATE", los_ate$aipw),
  row_from_fit("ATT", los_att$iptw), row_from_fit("ATT", los_att$aipw)
)
knitr::kable(format_combined(estimand_rows), align = "llrrrr")
```

The `estimand` argument changes which population IPTW and AIPW target.
Switching from the default ATE to ATT moves the `los` estimate from roughly
2.9 (IPTW) / 2.6 (AIPW) additional days to roughly 1.8 / 1.7 days — the
estimate for patients who received RHC rather than for the full analytic
sample. Matching (the second row of the main table) is ATT by construction but
reports roughly 3.1 days; that is not a contradiction. Matching analyzes only
the treated patients who found a control match within the 0.2-SD caliper — a
subset of the treated population — whereas the SMR weights retain every
treated unit, and the two can differ under effect modification. All three rows
target treated patients; exact agreement is not expected. The point is that
the parameter behaves as designed: it shifts the propensity-score methods to a
different, explicitly chosen population. Note that at this sample size the ATT
estimates are also noisier — their confidence intervals cross zero — so the
shift is informative about the estimand, not a statistically distinguishable
difference between ATE and ATT.

### Weight trimming

```{r los-trim}
los_trim <- fit_outcome(los_sample, exposure, covariates, "los",
                        type = "continuous", method = c("iptw", "aipw"),
                        trim = c(0.01, 0.99))

ci_str <- function(fit) {
  sprintf("%.2f (%.2f-%.2f)", fit$estimate, fit$conf_low, fit$conf_high)
}
trim_rows <- data.frame(
  Method        = c("IPTW", "AIPW"),
  "No trimming" = c(ci_str(los_ate$iptw), ci_str(los_ate$aipw)),
  "trim = c(0.01, 0.99)" = c(ci_str(los_trim$iptw), ci_str(los_trim$aipw)),
  check.names = FALSE
)
knitr::kable(trim_rows)
```

Truncating the 1% most extreme weights on each side of the distribution barely
moves the estimate (IPTW 2.88 to 2.91, AIPW 2.61 to 2.49 days) while modestly
narrowing the confidence intervals. That is the expected behavior on this
cohort: the positivity check above showed the weight extremity is real but
mild, so trimming dampens variance without pulling the answer somewhere new. In
a smaller or more imbalanced sample — where a handful of extreme weights can
dominate — the same `trim` argument is the lever for assessing how much the
headline estimate depends on those few units.

### Log-transformed `los`

```{r los-log}
los_sample$log_los <- log(los_sample$los)
los_log <- fit_outcome(los_sample, exposure, covariates, "log_los",
                       type = "continuous", method = "regression")
cat("Back-transformed effect of RHC on los:\n")
cat(sprintf("  exp(coef) = %.3f  (95%% CI %.3f-%.3f)\n",
            exp(los_log$estimate), exp(los_log$conf_low), exp(los_log$conf_high)))
```

This is the concrete version of Phase 18's `los` caveat: fitting the same
regression on the log-day scale back-transforms to a roughly 1.10 multiplicative
effect (95% CI comfortably above 1) — RHC is associated with roughly 5-17%
longer stay — versus the roughly +2.4-day additive effect on the original
scale. Same direction, same conclusion, and clear of the null, so the
day-scale headline in the main table is not an artifact of the outcome's
right-skew. The log-scale ratio is a sensitivity check, not a replacement for
the interpretable day-scale mean difference.

## Exporting results

```{r export, eval = FALSE}
for (i in seq_len(nrow(outcomes))) {
  nm <- outcomes$name[i]
  export_results(results_list[[nm]], output_dir = paste0("IndepAssoc_RHC_", nm))
}
write.csv(combined_comparison, "IndepAssoc_RHC_all_outcomes_comparison.csv",
          row.names = FALSE)
```

(Not run automatically when this vignette builds, to avoid writing files
outside the package's own directories during `R CMD check`.)

## Caveat

This is a confounder-adjusted *association* under the standard
no-unmeasured-confounding assumption — not a proven causal effect. Even five
methods agreeing in direction cannot rule out an unmeasured confounder that
happens to affect all five estimators the same way.

## References

- Connors, A.F., Speroff, T., Dawson, N.V., et al. (1996). The Effectiveness
  of Right Heart Catheterization in the Initial Care of Critically Ill
  Patients. *JAMA*, 276(11), 889-897.
- The RHC data ship with the package as `rhc_sample`; they originate from the
  SUPPORT study and are mirrored at https://hbiostat.org/data/repo/rhc.csv.
