---
title: "Validating the Methods on the RHC Benchmark Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating the Methods on the RHC Benchmark Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The simulated examples in the package vignette prove the pipeline runs; this
vignette validates the five confounding-adjustment methods on a real,
well-known observational dataset. The RHC (Right Heart Catheterization) study
collected data on 5735 critically ill adults enrolled in the SUPPORT trial;
Connors et al. (1996, *JAMA*) found that right heart catheterization was
associated with increased mortality and increased use of resources.

Our comparison against that literature is strictly directional: RHC is
associated with harm in the literature, and we check whether all five methods
agree in direction. It is not an exact-value match to any published point
estimate.

The binary exposure is `swang1` (1 = RHC). We analyze two outcomes with the
package's five methods — outcome regression, matching, stratification, IPTW,
and AIPW: `dth30`, the true 30-day mortality (binary), and `los`, the length of
stay in days (continuous).

```{r data}
rhc_ok <- FALSE
rhc_path <- NULL
candidates <- c(
  file.path("..", "rhc_data", "rhc.csv"), # vignettes/ working dir during build
  file.path("rhc_data", "rhc.csv"), # repo-root-relative, interactive
  "/home/mostafa/my-coding-project/rhc_data/rhc.csv" # developer machine
)
for (cand in candidates) {
  if (file.exists(cand)) {
    rhc_path <- cand
    break
  }
}
if (is.null(rhc_path)) {
  dest <- file.path(tempdir(), "rhc.csv")
  dl_ok <- tryCatch(
    suppressWarnings(download.file("https://hbiostat.org/data/repo/rhc.csv",
                                   dest, quiet = TRUE, mode = "wb")),
    error = function(e) 1L
  )
  if (identical(as.integer(dl_ok), 0L) && file.exists(dest)) {
    rhc_path <- dest
  }
}
if (is.null(rhc_path)) {
  message("The RHC data file could not be found locally and the download ",
          "from https://hbiostat.org/data/repo/rhc.csv failed; ",
          "the analysis chunks below are skipped.")
} else {
  rhc_ok <- TRUE
}
```

## Data preparation

```{r preprocessing, eval = rhc_ok}
library(IndepAssoc)

rhc <- prepare_rhc_data(rhc_path)
nrow(rhc$data)
length(rhc$covariates)
table(rhc$data$swang1)
sum(complete.cases(rhc$data[, rhc$covariates]))
```

`prepare_rhc_data()` returns a cleaned data frame with n = 5735 rows and a
48-column covariate vector. The exposure counts are 2184 RHC patients and 3551
controls, and complete-case analysis on the 48-covariate set retains the full
5735 rows: `adld3p` and `urin1` are excluded from the covariate set because
their missingness would collapse the sample. The raw CSV stores the `cat2`
category "None" as the literal text `"NA"`; the helper recodes that to `"None"`
before converting the remaining literal `"NA"` strings to real missing values.

Note the distinction between the two binary outcomes in the raw data: `death`
is a longer-term all-cause outcome, whereas `dth30` is the true 30-day
mortality. This vignette reports `dth30`.

## Binary outcome: 30-day mortality (dth30)

```{r binary-pipeline, eval = rhc_ok}
# We set a fixed random seed (1) before running the pipeline so the matching
# step is reproducible; see `?run_pipeline`.
set.seed(1)
res30 <- run_pipeline(rhc$data, "swang1", rhc$covariates, "dth30",
                      type = "binary")
res30$comparison
```

Here `estimate` is the adjusted odds ratio for 30-day mortality associated with
RHC. All five methods — outcome regression, matching, stratification, IPTW, and
AIPW — find RHC associated with *increased* 30-day mortality, with odds ratios
near 1.3 and confidence intervals that exclude 1. The convergence across five
functionally different adjustment strategies (those that model the outcome
directly and those that model treatment assignment) is directionally consistent
with Connors et al. (1996) and the later RHC reanalysis literature: the harm
direction is reproduced by every method. The comparison table is the primary
evidence for or against this independent association.

This comparison is directional and qualitative — RHC is associated with harm —
not an exact match to any specific published point estimate.

```{r binary-forest, eval = rhc_ok}
plot_comparison(res30$comparison, log_scale = TRUE)
```

## Continuous outcome: length of stay (los)

```{r continuous-pipeline, eval = rhc_ok}
# `los` (discharge date minus admission date) has one missing value; drop it
# before running the pipeline so the complete-case sample is n = 5734.
rhc_cc <- rhc$data[!is.na(rhc$data$los), ]
nrow(rhc_cc)

# We set a fixed random seed (1) before running the pipeline so the matching
# step is reproducible; see `?run_pipeline`.
set.seed(1)
reslos <- run_pipeline(rhc_cc, "swang1", rhc$covariates, "los",
                       type = "continuous")
reslos$comparison
```

Here `estimate` is the adjusted mean difference in length of stay (days)
associated with RHC. All five methods find RHC associated with a longer stay,
roughly +3 to +4 days, with confidence intervals that exclude 0. The outcome
`los` is derived as discharge date minus admission date and has a single
missing value; dropping that row leaves the complete-case sample of n = 5734
used here. As with the mortality analysis, this association is directional only
— it is not an exact match to any published point estimate.

```{r continuous-forest, eval = rhc_ok}
plot_comparison(reslos$comparison, log_scale = FALSE)
```

## Caveat

This is a confounder-adjusted association under the standard
no-unmeasured-confounding assumption, not a proven causal effect.

## References

- Connors, A.F., Speroff, T., Dawson, N.V., et al. (1996). The Effectiveness of
  Right Heart Catheterization in the Initial Care of Critically Ill Patients.
  *JAMA*, 276(11), pp.889-897.
- The RHC data are distributed with the `Hmisc` package's `rhc` dataset and
  mirrored at https://hbiostat.org/data/repo/rhc.csv.
